\textit{MSF: Modulated Sub-graph Finder}

### 2018-05-14 MSF_Paper_FA.tex
Figures with different Cut offs

To demonstrate its usefulness, \texttt{MSF} is applied to an RNA-seq data
set of primary human monocyte-derived macrophages (MDMs) infected with
Ebola virus (GSE84188)~\cite{Olejnik}. Ebola Virus (EBOV) belongs to the
Filoviridea family; filamentous, enveloped and single stranded RNA
viruses. EBOV causes hemorrhagic fever in humans, inducing the host innate
and adaptive immune response to be unable to control virus
infection~\cite{Prins}. Until now, there are no approved antiviral drugs
for the treatment of Ebola virus infection~\cite{Konde,Rhein}.  The initial
targets of EBOV are the macrophages and dendritic immune
cells~\cite{Falasca,Rhein}. EBOV inhibits the critical innate immune
response of the host, which includes the activation of alpha/beta
interferon (IFN-$\alpha / \beta$)~\cite{Prins,Konde,Cardenas}. It has been
proposed that IFN-$\alpha / \beta$ should be tested against Ebola for its
antiviral activity through clinical trials~\cite{Konde}. Ebola infection
data was selected to test the approach because it has been well recognized
in the literature for the last several decades \TODO{what has been
  recognized??}.  Thereby facilitating the verification of the results of
\texttt{MSF} with the vast literature present on Ebola
infection. Especially, the detection of IFN-$\alpha / \beta$ as point of
action for the virus, could be considered as an basic indicator of the
rightness and usefulness of the approach.

@@

The EBOV infection experiments describe the course of infection at three
time-points 6, 24 and 48 hour post infection (hpi). For the earliest time
point at 6~hpi, five modulated sub-graphs were identified with 41, 107, and
18 genes. Two of the five sub-graphs were less than 4 genes long. Most of
the genes part of the sub-graphs were cytokines, chemokines (CXCL10, CCL8)
and Interleukin genes (IL6, IL27, IL23) which serve as a subset \TODO{??}
of CD4+ T helper cells. IFNB1 and IFNA1 were both identified as two of the
possible sources. Most of the sources identified by \texttt{MSF} were
type~I interferon induced genes. At 24~hpi nine modulated sub-graphs were
identified with four main \TODO{(more than X genes)} sub-graphs consisting
of 38, 85, 148 and 167 genes. IFNA1 was identified as one of the sources in
the most significantly modulated sub-graph. For the last time-point 48~hpi,
eleven modulated sub-graphs were identified. Eight of the sub-graphs were
less than seven genes and main sub-graphs had 39, 71 and 210 genes. IFNB1
and IFNA1 were identified as the two sources out of the possible sources
\TODO{??}.

@@

 To assess the robustness and stability of our method, Poisson distributed
 noise was added to the read counts of the three time-points data set, used
 in the previous analysis. Then DGEA was carried out on the disturbed data
 with the same parameters as for the native data using edgeR, followed by
 analysis with \texttt{MSF}. This procedure was carried out 100 times.
 Every time the genes from the modulated sub-graphs identified from noisy
 data were compared to the genes of sub-graphs identified from the native
 data. This was also done for the DEG obtained for each run, a cutoff of
 FDR less than 0.05 was used to compare the retrieved DEG from noisy data
 to the native data.  The robustness of \texttt{MSF} and the DEG analysis
 for the time-point 6, 24, and 48~hpi are shown in
 Fig.~\ref{fig:DEGvsMSF}. The procedure how data noise was modeled can be
 considered as rather stringent, which is already reflected by the limited
 recall rate in the edgeR based DGEA, between 84~\% (6~hpi) and 93~\%
 (24~hpi). For \texttt{MSF}-analysis the observed median recall rates lay
 between 71~\% (6~hpi) and 84~\% (48~hpi). The better performance of the
 pure DGEA can be explained by the fact that disturbed p-values do not
 change the results for DGEA as long as the p-value does not wise above the
 chosen cutoff value. In contrast, \texttt{MSF} is sensitive to p-value
 changes across the whole range of possible values.

@@

Gene enrichment analysis was performed on genes in \texttt{MSF} identified
sub-graphs for each time-point using Reactome Analyze data
tool~\cite{Reactome}. Reactome's over-representation analysis tests whether
certain Reactome pathways are enriched by the list of genes submitted. For
comparison the DEG results from edgeR were filtered using a cut off of
adjusted p-value < 0.1. This subset of DEG list was used for gene
enrichment analysis. Enriched pathways were further filtered by confidence
score p-values, pathways having p-value less $<$ 0.05 for cut off DEG list
and $>$ 1$^{-10}$ for pathway enrichment in \texttt{MSF} identified
sub-graph genes. Out of 282 pathways dis-regulated in the two comparison
analysis with three time-points each, pathways that were present in at
least three time-points are shown in Fig~\ref{fig:msfvsreactome} \TODO{rephrase; not really clear how this filtering of pathways shown was done}.

The comparison shows most of the pathways known from literature
to be dis-regulated by Ebola infection are enriched in both the enrichment
analysis. Toll-Like receptor signaling pathway when interacts with EBOV
glycoprotein (GP), triggers the activation of
cytokines~\cite{Olejnik}. This pathway was not identified when a p-value
cut off DEG list was analyzed for enrichment. Due to \texttt{MSF}
considering the complete DEG results, even the weak signal at the earliest
time-point was detected; for example Toll-like receptor signaling. \TODO{what is the point you want to say and state it clearly. this is to vague}

@@

The Ebola infection RNA-seq dataset analyzed during the current study are
available in the GEO repository (GSE84188)~\cite{Olejnik}. The cell
signaling network file used is from Reactome Functional interactions (FIs)
Version 2016~\cite{Cytokegg}

@@

\begin{figure*}[p]
	\centering
	\fbox{
		\begin{minipage}{17 cm}
			\includegraphics[width=16cm]{DEGvsMSF2.png}
			\caption{\TODO{make plot smaller and consider to add DEG vor 0.1, 0.05 and 0.01 as you did before} The percentage of DEG analysis genes
                          and MSF identified sub-graph genes recall
                          rates for the three different time points of
                          EBOV infection data for 100 simulation where
                          Poisson distributed noise was added to the
                          experimental deduced reads per gene
                          counts.\label{fig:DEGvsMSF}}
			\label{label}
		\end{minipage}
	}
\end{figure*}

\begin{figure*}[p]
	\centering
	\fbox{
		\begin{minipage}{10 cm}
	\includegraphics[width=1.0\linewidth]{MSFvsReactome}
	\caption{The heatmap shows Reactome pathway enrichment for genes in \texttt{MSF} identifed sub-graphs for the three time-points and the Reactome pathway enrichment for subset of DEG with adjusted p-value cut off of < 0.1. Green color shows the pathway being enriched and red color shows pathway not found as enriched. \TODO{this is not really a heatmap (no quantitative value encoded in the color but only qualitative value); you might want to include different p-value cutof for the DGE; in your set up you have to comapre column 1 with column 4; column 2 with column 5, and so on; that is suboptimal; the time points are not chronological ordered.}}
	\label{fig:msfvsreactome}
\end{minipage}
}
\end{figure*}

>>>>

To demonstrate its usefulness, \texttt{MSF} is applied to an RNA-seq data
set of primary human monocyte-derived macrophages (MDMs) infected with
Ebola virus (GSE84188)~\cite{Olejnik}. Ebola Virus (EBOV) belongs to the
Filoviridea family; filamentous, enveloped and single stranded RNA
viruses. EBOV causes hemorrhagic fever in humans, inducing the host innate
and adaptive immune response to be unable to control virus
infection~\cite{Prins}. Until now, there are no approved antiviral drugs
for the treatment of Ebola virus infection~\cite{Konde,Rhein}.  The initial
targets of EBOV are the macrophages and dendritic immune
cells~\cite{Falasca,Rhein}. EBOV inhibits the critical innate immune
response of the host, which includes the activation of alpha/beta
interferon (IFN-$\alpha / \beta$)~\cite{Prins,Konde,Cardenas}. It has been
proposed that IFN-$\alpha / \beta$ should be tested against Ebola for its
antiviral activity through clinical trials~\cite{Konde}. Ebola infection
data was selected to test the approach because it has been well recognized for the last several decades, and vast literature is available for the pathogenesis of Ebola. Thereby facilitating the verification of the results of
\texttt{MSF} with the vast literature present on Ebola
infection. Especially, the detection of IFN-$\alpha / \beta$ as point of
action for the virus, could be considered as an basic indicator of the
rightness and usefulness of the approach.

@@

The EBOV infection experiments describe the course of infection at three
time-points 6, 24 and 48 hour post infection (hpi). For the earliest time
point at 6~hpi, five modulated sub-graphs were identified with 41, 107, and
18 genes. Two of the five sub-graphs were less than 4 genes long. Most of
the genes part of the sub-graphs were cytokines, chemokines (CXCL10, CCL8)
and Interleukin genes (IL6, IL27, IL23). IFNB1 and IFNA1 were both identified as two of the
possible sources. Most of the sources identified by \texttt{MSF} were
type~I interferon induced genes. At 24~hpi nine modulated sub-graphs were
identified with four main sub-graphs consisting
of 38, 85, 148 and 167 genes, others being smaller than 10 genes. IFNA1 was identified as one of the sources in
the most significantly modulated sub-graph. For the last time-point 48~hpi,
eleven modulated sub-graphs were identified. Eight of the sub-graphs were
less than seven genes and main sub-graphs had 39, 71 and 210 genes. IFNB1
and IFNA1 were identified as the two sources out of the total sources.

@@

 To assess the robustness and stability of our method, Poisson distributed
 noise was added to the read counts of the three time-points data set, used
 in the previous analysis. Then DGEA was carried out on the disturbed data
 with the same parameters as for the native data using edgeR, followed by
 analysis with \texttt{MSF}. This procedure was carried out 100 times.
 Every time the genes from the modulated sub-graphs identified from noisy
 data were compared to the genes of sub-graphs identified from the native
 data. This was also done for the DEG obtained for each run, using three different cutoffs of
 FDR 0.01, 0.05 and 0.1. The robustness of \texttt{MSF} and the DEG analysis
 for the time-point 6, 24, and 48~hpi are shown in
 Fig.~\ref{fig:DEGvsMSF}. The procedure how data noise was modeled can be
 considered as rather stringent, which is already reflected by the limited
 recall rate in the edgeR based DGEA, between 68~\% (6~hpi) and 93~\%
 (24~hpi). For \texttt{MSF}-analysis the observed median recall rates lay
 between 71~\% (6~hpi) and 84~\% (48~hpi). The better performance of the
 pure DGEA can be explained by the fact that disturbed p-values do not
 change the results for DGEA as long as the p-value does not wise above the
 chosen cutoff value. In contrast, \texttt{MSF} is sensitive to p-value
 changes across the whole range of possible values.

@@

Gene enrichment analysis was performed using Reactome Analyze data
tool~\cite{Reactome}. Reactome's over-representation analysis tests whether
certain Reactome pathways are enriched by the list of genes submitted.Genes from \texttt{MSF} identified
sub-graphs for each time-point were analyzed for gene enrichment analysis.  For
comparison the DEG results from edgeR were filtered using three different cut off of
adjusted p-value 0.01, 0.05 and 0.1. This three subsets of DEG list were used for gene
enrichment analysis. The compression of \texttt{MSF}  and the three subset DEG list is shown in Fig~\ref{fig:msfvsreactome}

The comparison shows most of the pathways known from literature
to be dis-regulated by Ebola infection are enriched in both the enrichment
analysis. Toll-Like receptor signaling pathway when interacts with EBOV
glycoprotein (GP), it triggers the activation of
cytokines~\cite{Olejnik}. Toll-like receptor pathway is expected to be dis-regulated in the early stage of infection,  this pathway was not identified as significantly dis-regulated when p-value
cut off DEG lists were analyzed for enrichment. Ten toll-like receptor cascades were identified as dis-regulated from gene enrichmant analysis of \texttt{MSF} identified genes, not a single one of these pathway was shown to be dis-regulated in DEG cut off lists. Since \texttt{MSF}
considers the complete DEG results, even the weak signal at the earliest
time-point was detected; for example Toll-like receptor signaling.

@@

Tlhe Ebola infection RNA-seq dataset analyzed during the current study are
available in the GEO repository (GSE84188)~\cite{Olejnik}. The cell
signaling network file used is from Reactome Functional interactions (FIs)
Version 2016~\cite{Cytokegg}

@@

\begin{figure*}[p]
	\centering
	\fbox{
		\begin{minipage}{12 cm}
			\includegraphics[width=12cm]{DEGvsMSF.png}
			\caption{Shows the percentage of DEG analysis (3 different cut-offs) genes
                          and MSF identified sub-graph genes recall
                          rates for the three different time points of
                          EBOV infection data for 100 simulation where
                          Poisson distributed noise was added to the
                          experimental deduced reads per gene
                          counts.\label{fig:DEGvsMSF}}
			\label{label}
		\end{minipage}
	}
\end{figure*}

\begin{figure*}[p]
	\centering
	\fbox{
		\begin{minipage}{10 cm}
	\includegraphics[width=1.0\linewidth]{MSFvsReactomeUpsetPlot}
	\caption{The Upset plot shows the number of common pathways between different time-point and cut offs. }
	\label{fig:msfvsreactome}
\end{minipage}
}
\end{figure*}
